Assessing landscape N removal in coastal New England catchments using the N-Sink

### 2022-03-14 manuscript/manuscript.Rmd
Version sent to Jeff, Art and Chet

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold, A.^*2*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features within watersheds can act as landscape
  sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous
  N, effectively removing it from the aquatic system. Identifying and evaluating these
  landscape sinks and their spatial relationship to N sources can assist managers
  to evaluate the effects of their decisions on downstream resources. N-Sink is an
  approach that uses widely available GIS data to identify landscape sinks within
  HUC-12 catchments, estimate their N removal potential and map the effect of those
  sinks on N movement through the catchment. A series of static maps are produced
  to visualize N removal efficiency, N loading, transport and delivery, the latter
  three in the form of indices, To facilitate data acquisition, processing and visualization,
  an R package `nsink` was developed and has been used to produce static maps for
  all HUC-12 catchments in southern New England. Users can use the R package to investigate
  a HUC-12 of interest, or can visit the UConn website to explore the already mapped
  areas. Users can also investigate specific flowpaths interactively by clicking on
  any location within the catchment. A flowpath is generated, along with a table describing
  N removal along each segment.
---

@@

```{r setup, echo=FALSE, warning = FALSE}

library(nsink)
```

@@

## Development of the R package `nsink`
The package `nsink` implements the approach detailed in Kellogg et al. [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. That version required time-consuming data manipulation by hand due to limitations of earlier NHD datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink could be automated and more easily applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`.

@@

```{r static_maps, eval=TRUE, include = FALSE}
niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 900)
```

@@

Use Examples
============
### Example 1: Comparison of N delivery from two adjacent locations
Comparing the N delivery from two locations that are separated by approximately 900 meters illustrates the importance of location with respect to landscape N sinks. Approximately 56% of the N that originates at point A is transported to the estuary vs. approximately 25% of the N that originates at point B (Fig 4). The difference is due to the different N sinks encountered along the way to the outlet or estuary. Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2). Flowpath B encounters two ponds along the way, significantly increasing residence time and N transformation (Table 3).

```{r plots4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates. Hydrofeatures are pictured on the NTI (Nitrongen Transport Index) map."}

@@

niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
                     "#F9C159", "#F99844", "#F56A2E","#EF2820"),
                 main = "Flowpaths originating at points A and B, separated by ~ 900 m")
plot(niantic_data$streams, add = TRUE, col = "dodgerblue")
plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray50", lwd = 3.0)
plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray50", lwd = 3.0)
plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
x <- c(pta[1], ptb[1])
y <- c(pta[2], ptb[2])
ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
ab_plot <- st_sf(st_sfc(ab), crs = aea)
plot(ab_plot, add = TRUE, pch = 16)
text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)

```
```{r compare_flowpaths, include = TRUE, echo = FALSE}
knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
```

### Example 2: Source controls and sink protection or restoration
The maps produced by nsink can be helpful in prioritizing source controls, sink protection and/or restoration. While wetlands are considered "protected" by law they are frequently altered through variances granted by local commissions or committees charged with zoning decisions. By better understanding the role sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving water. Variances granted on a case=by-case basis can, however unintentionally, quickly undermine local water quality goals.

Similarly, sources of N can be highlighted using the nsink static maps [Fig XX]. The N Delivery Index map combines N sources with N transport to highlight

Potential ways to use N-Sink are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.

Discussion and conclusions
==========================

Really great conclusions and discussion thereof goes here!

>>>>

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold, A.^*2*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features within watersheds can act as landscape
  sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous
  N, effectively removing it from the aquatic system. Identifying and evaluating these
  landscape sinks and their spatial relationship to N sources can assist managers
  to evaluate the effects of their decisions on downstream resources. N-Sink is an
  approach that uses widely available GIS data to identify landscape sinks within
  HUC-12 catchments, estimate their N removal potential and map the effect of those
  sinks on N movement through the catchment. A series of static maps are produced
  to visualize N removal efficiency, N loading, transport and delivery, the latter
  three in the form of indices, To facilitate data acquisition, processing and visualization,
  an R package `nsink` was developed and has been used to produce static maps for
  all HUC-12 catchments in southern New England. Users can use the R package to investigate
  a HUC-12 of interest, or can visit the UConn website to explore the already mapped
  areas. Users can also investigate specific flowpaths interactively by clicking on
  any location within the catchment. A flowpath is generated, along with a table describing
  N removal along each segment.
---

@@

```{r setup, echo=FALSE, warning = FALSE}
library(FedData)
library(nsink)
library(raster)
```

@@

## Development of the R package `nsink`
The package `nsink` implements the approach detailed in Kellogg et al. [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. That version required time-consuming data manipulation by hand due to limitations of earlier NHD datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink could be automated and more easily applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`. The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].

@@

```{r static_maps, eval=TRUE, include = FALSE}
niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 450)
```

@@

Use Examples
============
### Example 1: Comparison of N delivery from two adjacent locations
Users can use the following to interactively identify a starting point for a flowpath of interest. 
```{r case1, echo = TRUE, eval = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072
# Select location on map for starting point
pt <- unlist(locator(n = 1))
# Convert to sf POINT
start_loc <- st_sf(st_sfc(st_point(pt), crs = aea))
# Generate the flowpath
niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
# Summarize N removal along the flowpath
niantic_removal <- nsink_calc_removal(niantic_data)
niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) # table of removal within flowpath segments
```

Using this approach, we can compare the N delivery from two locations that are separated by approximately 900 meters. This example illustrates the importance of location with respect to landscape N sinks. Approximately 56% of the N that originates at point A is transported to the estuary vs. approximately 25% of the N that originates at point B (Figure 4). The difference is due to the different N sinks encountered along the way to the estuary. Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2). Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3). These two undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet. If development were proposed in the western portion of this area, source controls could be advised, while the less "leaky "eastern region underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing N transformation.

```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates. Hydrofeatures are pictured on the NTI (Nitrongen Transport Index) map."}

@@

niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
                     "#F9C159", "#F99844", "#F56A2E","#EF2820"),
                 main = "Flowpaths originating at points A and B, separated by ~ 900 m")
plot(niantic_data$streams, add = TRUE, col = "dodgerblue")
plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray50", lwd = 3.0)
plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray50", lwd = 3.0)
plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
x <- c(pta[1], ptb[1])
y <- c(pta[2], ptb[2])
ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
ab_plot <- st_sf(st_sfc(ab), crs = aea)
plot(ab_plot, add = TRUE, pch = 16)
text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
```

```{r compare_flowpaths, include = TRUE, echo = FALSE}
knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
```

### Example 2: Source controls and sink protection or restoration
The maps produced by N-sink can be helpful in prioritizing source controls, sink protection and/or restoration. For example, wetlands are considered "protected" by law but are frequently altered through variances granted by local commissions or committees charged with zoning decisions. By better understanding the role N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters. Variances granted on a case=by-case basis can quickly undermine local water quality goals. Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize acquisition or wetland restoration projects.

Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps. The N Delivery Index (NDI) map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment. Figure 5A shows land cover in a smaller area of the watershed, with two developed areas circled. Figure 5B shows the N Delivery Index (NDI) from these two developed areas. Note that the northern area is less densely developed than the southern area but the NDI is higher in the northern area as compared to the southern area, suggesting that N source controls within the northern development would be worth investigating.

```{r figure5a, include = TRUE, echo = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest, highlighting two developed areas."}
box <- c(1949902, 1953930, 2288202, 2292262)
e <- extent(box)
legend_nlcd <- pal_nlcd()
case2 <- crop(niantic_data$nlcd, e)
vals <- unique(case2@data@values)
leg <- legend_nlcd[legend_nlcd$code %in% vals,]
case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
case2_south <- st_sf(st_sfc(st_point(c(1952693, 2289249)), crs = aea))

my_key <- list(rectangles=list(col = leg$color),
              text=list(lab=leg$description),
              space='left',
              columns=1,
              size=2,
              cex=.6)
plot(case2, col = legend_nlcd$color)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
```
```{r, figure5b, include = TRUE, echo = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index (NDI) over land cover, with hotter colors denoting higher N delivery as estimated based on normalized loading from different land covers multiplied by the N Transport Efficiency (NTE)."}
case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
plot(case2, col = legend_nlcd$color, alpha = 0.6, legend = FALSE)
plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)

```

Other examples of N-Sink are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.

Conclusions
==========================

N-Sink as implemented in the `nsink` R package allows users to easily download all necessary spatial data (hydrography, land cover and soils), create static maps (N Removal Efficiency, N Transport Efficiency and N Delivery Index), and interactively investigate N removal along flowpaths within any specified catchment. Currently, the default catchment definition uses HUC-12 designations, though it is also possible to use larger catchments. Larger catchments would require longer download and processing times. For investigators who do not work in R, we have created an interactive web application that covers HUC-12 catchments in coastal CT, RI and southern MA [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).

Local decision-makers that have prioritized nitrogen mitigation in their long-term planning efforts can use this tool to help them better understand the potential impact of proposed development projects and zoning variances. Similarly, land trusts and other NGOs interested in N mitigation can use this tool to identify high priority areas for acquisition or restoration.
